smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies

### 2023-08-22 manuscript/manuscript.qmd
modified language on group 2 diagnostic model; abbreviate missingness mechanisms in box

-   **MCAR**: The missingness does not depend on any other observed or unobserved covariate(s).

-   **MAR**: The missingness depends and can be explained by other observed covariates.

-   **MNAR**: The missingness depends on unoberserved covariate(s). For example, the missingness may be explained by other covariate(s) which is/are not observed in the underlying dataset (MNAR~unmeasured~). The missingness can also just dependent on the actual value of the partially observed covariate itself (MNAR~value~).
:::

@@

The workflow is generally categorized into descriptives, pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and how these results can be interpreted. Examples are illustrated using a simulated EHR dataset that is part of the package (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would indicate a random prediction and translate to a potential MCAR or MNAR mechanism.

@@

The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**). The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.

@@

Missing data are ubiquitous in EHR and may lead to bias if not handled appropriately. We developed `smdi` to streamline routine diagnostics of missing data. By cross-checking the resulting diagnostic parameters to expected estimates (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness patterns and mechanisms and help elucidate if analytic approaches such as imputation analyses are viable options.

>>>>

-   **Missing completely at random (MCAR)**: The missingness does not depend on any other observed or unobserved covariate(s).

-   **Missing at random (MAR)**: The missingness depends and can be explained by other observed covariates.

-   **Missing not at random (MNAR)**: The missingness depends on unoberserved covariate(s). For example, the missingness may be explained by other covariate(s) which is/are not observed in the underlying dataset (MNAR~unmeasured~). The missingness can also just dependent on the actual value of the partially observed covariate itself (MNAR~value~).
:::

@@

The workflow is generally categorized into descriptives, pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and how these results can be interpreted. Examples are illustrated using a synthetic EHR dataset that is part of the package and simulates an oncology cohort with a binary exposure, a time-to-event outcome and several confounders and prognostic covariates including three POCs (EGFR and PD-L1 \[biomarkers\] and ECOG \[performance score\]) following MCAR, MAR and MNAR mechanisms (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>).

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of each POC given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would indicate the model is unable to discriminate missing vs. observed values based on available data; this could be due to a mechanism that is close to MCAR or one where the missingness is associated with unmeasured data (MNAR).

@@

The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**). By cross-checking the resulting diagnostic parameters to expected estimates (@fig-results, [@weberpals2023]), the diagnostics can provide valuable insights into underlying missingness mechanisms and thereby help elucidate if analytic approaches such as imputation analyses are viable options.

The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.

@@

Missing data are ubiquitous in EHR and may lead to bias if not handled appropriately. We developed `smdi` to streamline routine diagnostics of missing data.
